---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# assign4nspack

<!-- badges: start -->
<!-- badges: end -->

## Introduction

**assign4nspack** is a teaching-focused R package for exploring **high-frequency freshwater water-quality and nitrate relationships** using data derived from the **National Ecological Observatory Network (NEON)**.

It provides:

- Cleaned, curated hourly datasets (`wq_hourly`, `nitrate_hourly`, `wq_nitrate`)
- Data-processing scripts in `data-raw/`
- An interactive **Shiny app** to visualise water-quality and nitrate trends across sites  
- Full documentation and reproducible examples built with **roxygen2** and **pkgdown**

## Installation

You can install the development version of assign4nspack from [GitHub](https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-NutchanokS) with:

``` r
# install.packages("pak")
pak::pak("ETC5523-2025/assignment-4-packages-and-shiny-apps-NutchanokS")
```

Then load the package:

``` r
library(assign4nspack)
```

## Datasets Included

The package ships with three tidy datasets prepared from NEON sensor data:

| Dataset | Description |
|----------|-------------|
| **`wq_hourly`** | Hourly water-quality data (conductivity, dissolved oxygen, turbidity) from NEON **DP1.20288.001** |
| **`nitrate_hourly`** | Hourly nitrate concentrations (µmol/L) from NEON **DP1.20033.001** |
| **`wq_nitrate`** | Combined dataset aligning water-quality and nitrate observations by site, sensor, and timestamp |

### Data filtering and coverage

Data were sourced from NEON freshwater sites covering the period **January 2018 to December 2019**.  
For each site, both upstream (`S1_upstream`) and downstream (`S2_downstream`) sensors were used when available.  
Raw 15-minute observations from NEON products DP1.20288.001 (Water Quality) and DP1.20033.001 (Nitrate)  
were aggregated to **hourly means**, quality-filtered (`QF == 0`), and merged by `site`, `sensor`, and `date_time`.

The final datasets include hourly records for **multiple NEON stream sites** consistent with  
the study period described in *Kermorvant et al. (2023):*  
> *“Understanding links between water-quality variables and nitrate concentration in freshwater streams using high-frequency sensor data.”*

This ensures the datasets reflect comparable temporal coverage and data quality to the published analysis.

### Example preview

```{r example-data, message=FALSE}
library(dplyr)
data("wq_nitrate", package = "assign4nspack")
glimpse(wq_nitrate)
```

## Key variables

- `site` — 4-letter NEON site identifier (e.g., ARIK, LEWI)
- `sensor` — sensor position (S1_upstream, S2_downstream)
- `date_time` — hourly timestamp (UTC)
- `cond_uScm`, `do_mgL`, `turb_FNU` — water-quality variables
- `nitrate_umolL` — nitrate concentration (µmol/L)

## Shiny App

The package includes an interactive Shiny dashboard to explore how nitrate concentrations relate to water-quality measures across NEON sites.

Launch it with:

``` r
assign4nspack::run_app()
```
### App overview

| Tab | Purpose |
|----------|-------------|
| **About the Data** | Explains NEON products, variable meanings, and interpretation guidance|
| **Time Trend** | Line plots showing changes in conductivity, oxygen, turbidity, or nitrate over time |
| **Boxplot by Site** | Compare distributions across sites |
| **Nitrate Relationship** | Explore nitrate vs. water-quality variables with optional LM/LOESS fits, correlation, and site facets |

---screenshot app

## Data Sources

All datasets are derived from publicly available NEON sensor products:

- **DP1.20288.001** — Aquatic Instrumentation: Water Quality (WQ)
- **DP1.20033.001** — Aquatic Instrumentation: Nitrate in Surface Water (NSW)

Data were aggregated to hourly resolution, quality-checked, and merged by site and sensor.
Used here solely for educational purposes in ETC5523 — Communicating with Data.

## Documentation & Website

**Package site:**
<https://etc5523-2025.github.io/assignment-4-packages-and-shiny-apps-NutchanokS/>


## License

This package is released under the **MIT License.**
See the included `LICENSE` file for details.

## Author

Nutchanok Saitin (Author & Maintainer)

## Acknowledgements

- Package infrastructure: `usethis`, `devtools`, `roxygen2`, `pkgdown`
- Data wrangling: `dplyr`, `tidyr`, `lubridate`, `purrr`
- Visualisation: `ggplot2`, `plotly`, `shiny`, `bslib`
- Original data source: *National Ecological Observatory Network (NEON)*
